Using best practices for forecasting biodiversity changes in breeding birds

### 2017-07-06 manuscript/draft.Rmd
Edit and add refs to Introduction

---
output:
  pdf_document: default
  html_document: default
---
# Moving towards large-scale data-intensive forecasting of species richness in breeding birds
# Using best practices for forecasting biodiversity changes in breeding birds
Authors: Harris, Taylor and White?

@@

Forecasting the future state of ecological systems is increasingly important for
planning, management, and evaluating how well ecological models capture the key
ecological processes governing a system [@clark2001, @dietz2017]. One area of
particular interest for forecasting is biodiversity since it is important for
ecosystem function, central to conservation planning, and expected to be
influenced by anthropogenic global change. Anticipating potential changes in
biodiversity is crucial for large scale management and conservation, and for
addressing debates regarding whether local scale diversity is declining and may
continue to decline in response to anthropogenic influences [@dornelas2014].

Previous efforts to predict future patterns of species richness, and diversity
more generally, have focused primarily on building species distributions models
to quantify the spatial relationships between the occurrence of individual
species and the environment. Forecasts for future environmental conditions are
then used to predict where each species will occur in the future and the
probabilities of occurrence are summed across species to predict future species
richness [e.g., @thomas2004]. Alternatively, models that directly relate spatial
patterns of species richness to the environment have been developed and
generally perform equivalently to species distribution modeling based methods
[@algar2009, @distler2015].

However, despite the emerging interest in forecasting species richness and other
aspects of biodiversity, little is known about how effectively we can anticipate
their dynamics. This is due in part to the long time scales over which many
ecological forecasts are applied and the resulting difficulty in assessing
whether the predicted changes occurred. What we do know comes from a small
number of hindcasting studies, where models are built using data on species
occurrence and richness from the past and evaluated on their ability to predict
contemporary patterns [e.g., @algar2009, @distler2015]. These studies are a
valuable first step, but lack several components that are important for
developing good forecasting models and understanding how accurately these models
can predict the future. These components of good forecasting and evaluation (Box
1) broadly involve: 1) expanding the use of data to include biological and
environmental time-series [@treddenick2016]; 2) accounting for uncertainty in
observations, processes, model choice, and forecast evaluation
[@clark2001, @dietz2017]; and 3) conducting meaningful evaluations of the
forecasts by hindcasting, archiving short-term forecasts, and comparing
forecasts to baselines to determine if the forecasts are more accurate than
assuming the system is basically static [@Perretti2013].

In this paper we attempt to forecast the species richness of breeding birds at
hundreds of locations throughout North America while following the best
practices in Box 1 for ecological forecasting. To do this we combine 30 years of
time-series data on bird distributions with monthly time-series of climate data
and satellite-based remote-sensing. This 30 year time-scale has only recently
become possible for large-scale time-series based forecasting and allows us to
model and assess changes a decade or more into the future in the presence of
shifts in environmental conditions on par with predicted climate change. We
compare traditional distribution modeling based approaches to spatial models of
species richness, time-series methods, and simple baselines. All forecasts model
uncertainty and observation, are evaluated across different time lags using
hindcasting, and are publicly archived to allow future assessment. We discuss
the implications of these practices for our understanding of, and confidence in,
the resulting forecasts, and how we can continue to improve on these approaches
in the future.

>>>>

---
title: Using best practices for forecasting biodiversity changes in breeding birds
author:
- David J. Harris
- Shawn D. Taylor
- Ethan P. White
output:
  pdf_document: default
  html_document: default
---

@@

Forecasting the future state of ecological systems is increasingly important for
planning, management, and evaluating how well ecological models capture the key
ecological processes governing a system [@clark2001; @houlahan2017;
@dietz2017]. One area of particular interest for forecasting is biodiversity
since it is important for ecosystem function, central to conservation planning,
and expected to be influenced by anthropogenic global change [@cardinale2012;
@diaz2015; @tilman2017]. Anticipating potential changes in biodiversity is
crucial for large scale management and conservation, and for addressing debates
regarding whether local scale diversity is declining and may continue to decline
in response to anthropogenic influences [@diaz2015; @dornelas2014].

Previous efforts to predict future patterns of species richness, and diversity
more generally, have focused primarily on building species distributions models
to quantify the spatial relationships between the occurrence of individual
species and the environment [@thomas2004; @urban2015]. Forecasts
for future environmental conditions are then used to predict where each species
will occur in the future and the probabilities of occurrence are summed across
species to predict future species richness [e.g., @thuiller2005]. Alternatively,
models that directly relate spatial patterns of species richness to the
environment have been developed and generally perform equivalently to species
distribution modeling based methods [@algar2009; @distler2015].

However, despite the emerging interest in forecasting species richness and other
aspects of biodiversity, little is known about how effectively we can anticipate
their dynamics. This is due in part to the long time scales over which many
ecological forecasts are applied and the resulting difficulty in assessing
whether the predicted changes occurred [@dietzinreview]. What we do know comes
from a small number of hindcasting studies, where models are built using data on
species occurrence and richness from the past and evaluated on their ability to
predict contemporary patterns [e.g., @algar2009; @distler2015]. These studies
are a valuable first step, but lack several components that are important for
developing good forecasting models and understanding how accurately these models
can predict the future. These components of good forecasting and evaluation (Box
1) broadly involve: 1) expanding the use of data to include biological and
environmental time-series [@treddenick2016]; 2) accounting for uncertainty in
observations, processes, model choice, and forecast evaluation
[@clark2001, @dietz2017]; and 3) conducting meaningful evaluations of the
forecasts by hindcasting, archiving short-term forecasts, and comparing
forecasts to baselines to determine if the forecasts are more accurate than
assuming the system is basically static [@Perretti2013].

In this paper we attempt to forecast the species richness of breeding birds at
hundreds of locations throughout North America while following the best
practices in Box 1 for ecological forecasting. To do this we combine 30 years of
time-series data on bird distributions with monthly time-series of climate data
and satellite-based remote-sensing. This 30 year annual time-scale has only
recently become possible for large-scale time-series based forecasting and
allows us to model and assess changes a decade or more into the future in the
presence of shifts in environmental conditions on par with predicted climate
change. We compare traditional distribution modeling based approaches to spatial
models of species richness, time-series methods, and simple baselines. All
forecasts model uncertainty and observation, are evaluated across different time
lags using hindcasting, and are publicly archived to allow future assessment. We
discuss the implications of these practices for our understanding of, and
confidence in, the resulting forecasts, and how we can continue to build on
these approaches improve ecological forecasting in the future.
